Novel genetic control of migratory diapause in Australian Monarch Butterflies

### 2023-04-13 paper/paper_draft.Rmd
more paper edits

Interestingly, however, there a population of monarchs in Australia in which migratory behavior has apparently re-emerged after a it was previously lost. This population, which was established sometime before 1871 [@Zalucki2004], is the product of a series of sequential introductions that crossed the Pacific Ocean from North America beginning sometime in the early 1840s after the introduction of their host plants enabled individuals to survive after being blown from the mainland [@Zalucki2004; @Pierce2014; @Hemstrom2022a]. Monarchs are not known to be migratory on any Pacific islands, and no migratory behavior was observed in Australia until the 1930s, when non-reproductive overwintering aggregations were observed in northern New South Wales [NSW, see @James2019]. These and other individuals were later shown to migrate hundreds of kilometers each year to their wintering grounds from more southerly, inland locations in NSW [@James2019;@jamesInductionReproductiveDormancy1983]. After over 90 years (and hundreds of generations) without any apparent migration, the behavior therefore re-emerged in these populations. Understanding how this happened could provide crucial context around the "extinction" of migration in monarchs and other species.

@@

@@

# Methods
## Background
Monarch butterflies are native to the Carribean islands; North, Central, and northern South America; and are an established but non-native in Europe, Africa, Australia, and on a range of Pacific Islands [@oplerFieldGuideWestern1999; @Zhan2014]. In North America, the species is well known for it's unique, multi-generational migratory life history wherein individuals dispersing northward in the spring do so over three to four generations and then return to their wintering grounds in a single step [@malcolmSpringRecolonizationEastern1993; @cockrellTimeTemperatureLatitudinal1993]. These wintering grounds are located either in central Mexico or the coast of California for those individuals which summer to the east or west of the Rocky mountains, respectively [@dingleDistributionMonarchButterfly2005;@urquhartAutumnalMigrationRoutes1978], although a handful also over winter in southern Florida and the Caribbean [@dockxDirectionalStabilizingSelection2007a; @dockxNORTHAMERICANMONARCH2004]. Despite summering and wintering in different locations, gene flow between the eastern and western monarch populations is apparently high enough to render them genetically homogeneous [@Hemstrom2022a; @tallaGenomicEvidenceGene2020;@lyonsLackGeneticDifferentiation2012].

In Southern Florida, the Caribbean, Central America, South America, and throughout most of their introduced range, however, monarchs are non-migratory [@dockxDirectionalStabilizingSelection2007a;@altizerPOPULATIONSMONARCHBUTTERFLIES2010;@northMonarchButterflyBiology2004]. Residency in these populations appears to be a derived trait which has arisen multiple times after the species' split from *Danaus erripus*, their closest extant relative [@Zhan2014]. The Pacific expansion of monarchs is relatively recent, with historical records indicating that they established first in Hawaii in the 1840s, likely as a result of individuals blown in during storm events that were able to survive on recently introduced milkweeds [@Zalucki2004; @Pierce2014]. Genetic evidence is consistent with an introduction to Hawaii during this time period [@Hemstrom2022a]. As described above, they then spread across the Pacific and reached Australia in approximately 1871 [@Zalucki2004], and were recorded migrating again in Australia sometime in the 1930s [see Figure 1; @James2019].

The North American migratory populations are the subject of wide-spread research on topics including but not limited to navigation and orientation [@Guerra2014; @Merlin1700; @Mouritsen10162], parasite ecology [@satterfieldMigratoryMonarchsThat2018], and life-history associated phenotypic variation [@Herman2001; @Zhan2014; @Zhan2011; @Dingle1977], and are differentiated from non-migratory individuals by a range of phenotypic variables such as reproductive diapause, wing shape, and other physiological factors [@altizerPOPULATIONSMONARCHBUTTERFLIES2010; @hermanSTUDIESADULTREPRODUCTIVE1981; @browerFuelingFallMigration2006]. Note, however, that the monarch migratory syndrome is not necessarily genetically linked and can become split apart by changes in selection pressures [@Tenger-Trolander2019]. The migratory populations outside of North America, however, are not yet heavily studied, and the mechanisms enabling migration in them are not yet well understood.

@@

We assessed reproductive diapause by determining the degree of oocyte development in each of our adult female butterflies. We observed that individuals fell into two clearly defined bins: `r sum(old.data$Phenotype == "C", na.rm = TRUE)` out of `r nrow(old.data)` had fully developed, chorionated eggs (visible ridges along the exterior of the egg), while `r sum(old.data$Phenotype != "C", na.rm = TRUE)` females did not. Of these latter individuals, a few (`r sum(old.data$Phenotype == "I", na.rm = TRUE) + sum(old.data$Phenotype == "N", na.rm = TRUE)`) had no yolk in their eggs whatsoever, but most had some degree of yolking. Since vitellogenesis generally only begins to a substantial degree after eclosion in monarchs [@Pan1976], we called any individuals with more than a small amount of yolking in their eggs as reproductively mature (N = `r sum(sample.meta$likely_pheno[sample.meta$year == 2018] == "M", na.rm = TRUE)` and `r sum(sample.meta$likely_pheno[sample.meta$year == 2018] == "R", na.rm = TRUE)` for reproductively immature and mature individuals, respectively, images of egg development in each individual are available from the authors upon request). Wings and bodies from all individuals were preserved in dry coin envelops or 95% ethanol, respectively. We note that while we refer to the postponement of reproductive development under other suitable conditions as diapause, other authors refer to this process in monarchs as oligopause [@jamesOVARIANDORMANCYDANAUS1982] or quiescence, since diapause implies an extended refractory period.

@@

In order to determine if diapause is part of a generalized migratory syndrome in Australia, we also briefly re-analyzed the connection between reproductive status and wing morphology in our 2016 samples [@Freedman2018]. In order to determine if wing shape or size was correlated with reproductive status, we fit a pair of linear mixed effect models with either with wing shape or size as response variables, the number of yolked oocytes as a fixed effect, and maternal family as a random effect using the R package nnlme [@Pinheiro2007].

@@

Genetic diversity levels were extremely similar between our two years ($\pi =$ `r round(diversity$weighted_mean_pi[2], 4)` and `r round(diversity$weighted_mean_pi[3], 4)` and $H_{o}$ = `r round(diversity$weighted_mean_ho[2], 4)` and `r round(diversity$weighted_mean_ho[3], 4)` for 2016 and 2018, respectively). Each year had a slight heterozygote excess ($F_{IS}$ = `r round(diversity$weighted_mean_fis[2], 3)` and `r round(diversity$weighted_mean_fis[3],3)` for 2016 and 2018, respectively), although that excess was lower if all samples were considered jointly ($F_{IS}$ = `r round(diversity$weighted_mean_fis[1], 3)`). Pairwise $F_{ST}$ between the two years was small but marginally significant (`r round(fst$weighted_mean_fst[1], 3)`, $p =$ `r round(fst$weighted_mean_fst_p[1], 3)`). We found no significant genetic divergence between the two sample years on either the PCA or the NGSadmix results, with clustering instead driven by family, although we did find that three of our 2018 maternal families were substantially different from all other samples on our first two PC axes (Figure 2).

@@

Diapause status was strongly biased by maternal family (Fisher's Exact Test, $p < 0.001 $, see Figure 3a). We identified one small region of chromosome 11 (between positions ~5,589,000 to ~5,620,000) containing many loci associated with diapause status well beyond the background level of association, including three SNPs from two different RAD-tags that had a $p$-value  $\le 5 \times 10^{-5}$ (Figure 3b). This region overlaps the gene DPOGS208560 (LOC116768596), an E3 ubiquitin ligase, and was still strongly associated with diapause if only the 2018 butterflies alone were examined. We also identified one locus with a $p$-value  $\le 5 \times 10^{-5}$ on chromosome 16, but there were no other significantly associated loci nearby.

@@

@@

It is difficult to say what these results entail for the conservation of migratory species as a whole. While monarchs may be able to quickly recover from the loss of migratory behavior should conditions improve, it is unlikely that this is the case for most other migratory species of conservation concern, particularly for vertebrates. For example, if we assume the latter hypothesis (migratory variance in Australian monarchs is the result of a new mutation), it is clear that monarchs have had ample evolutionary opportunity to re-acquire migratory mutations: monarchs have a generation time of approximately seven generations per year if continuously breeding, which means that monarchs in Australia went through roughly ~ `r 7*50` generations in the 50 or so years since they were first reported on the continent, and although they likely experienced a strong bottleneck initially, they were probably at a relatively large effective size for most of that time. Since the rate at which new mutations appear in a population is proportional to both effective populations size and generation time, monarchs have had a large opportunity space for the generation of new migratory alleles.

Alternatively, it is safe to assume that if migratory variance in Australian monarchs was maintained from North America, it is clear that this would only have been possible because selection was not actively acting against that variance. This is actually quite likely: monarchs in the Pacific are not exposed to substantial annual changes in day length or average temperature, and thus do not receive their main migratory cues. Environmentally triggered migratory variation, therefore, is probably selectively nearly neutral [although more fixed migratory variance, such as wing morphology, was likely selected against, driving the observed, repeated decreases in wing size and length observed in newly non-migratory monarch populations, see @Freedman2020], and is therefore more likely to be maintained in non-migratory populations.

In contrast, take for example Chinook and steelhead salmon (*Oncorhynchus tshawytscha* and *mykiss*, respectively). Pre-mature migration, a migratory life history variant in both species where some portion of the population return to their natal streams earlier than others, seems to be controlled by an ancestrally derived (but independently evolved in each species) allelic variant at the GREB1L gene, which has been lost from the majority of populations due to hydrological changes such as dam installation [@Princee1603198; @Thompson177]. Salmon have *much* longer generation times and smaller effective population sizes (several years in both species), and so we should not expect pre-mature migration to as quickly re-evolve in salmon as it did migration in monarchs. In addition, since the pre-mature migration phenotypes have been selected against due to anthropogenic influences [@Thompson177], GREB1L is unlikely to be selectively neutral in most populations and is therefore not likely to be maintained over time as standing genetic variance.

Nonetheless, our findings are still optimistic for the future of migration in monarch butterflies or wherever selection on migratory alleles is merely relaxed and not becoming increasingly negative.

>>>>

Interestingly, however, there a population of monarchs in Australia in which migratory behavior has apparently re-emerged after a it was previously lost. This population, which was established sometime before 1871 [@Zalucki2004], is the product of a series of sequential introductions that crossed the Pacific Ocean from North America beginning sometime in the early 1840s after the introduction of their host plants enabled individuals to survive after being blown from the mainland [@Zalucki2004; @Pierce2014; @Hemstrom2022a]. Monarchs are not known to be migratory on any Pacific islands, and no migratory behavior was observed in Australia until the 1930s, when non-reproductive overwintering aggregations were observed in northern New South Wales [NSW, see @James2019]. These and other individuals were later shown to migrate hundreds of kilometers each year to their wintering grounds from more southerly, inland locations in NSW [@James2019;@jamesInductionReproductiveDormancy1983]. After over 90 years (and hundreds of generations) without any apparent migration, the behavior re-emerged in these populations. Understanding how this happened could provide crucial context around the "extinction" of migration in monarchs and other species.

@@

Diapause control is also an phenotype of interest because it highlights a theoretical dichotomy in the types of genes that are associated with migratory life histories: migratory genes can either be directly causal, proximate migratory genes that are required prior for migration to occur (such as genes that control diapause, directional motion, and navigation), or facultative migratory genes that increase fitness in already migratory populations but are nonetheless not strictly required for migratory behavior to occur (such as genes that change wing or limb shapes, respiration, lipid storage, and muscle strength). For a newly migratory population, like monarchs in Queensland, we would expect to the strongest genetic associations with migration to occur in proximate migratory genes that cause traits like diapause rather than facultative genes which may take longer to increase in frequency following the evolution of migratory behavior.

@@

# Methods
## Background
Monarch butterflies are native to the Carribean islands; North, Central, and northern South America; and are an established but non-native in Europe, Africa, Australia, and on a range of Pacific Islands [@oplerFieldGuideWestern1999; @Zhan2014]. In North America, the species is well known for it's unique, multi-generational migratory life history wherein individuals dispersing northward in the spring do so over three to four generations and then return to their wintering grounds in a single step [@malcolmSpringRecolonizationEastern1993; @cockrellTimeTemperatureLatitudinal1993]. These wintering grounds are located either in central Mexico or the coast of California for those individuals which summer to the east or west of the Rocky mountains, respectively [@dingleDistributionMonarchButterfly2005;@urquhartAutumnalMigrationRoutes1978], although a handful also over winter in southern Florida and the Caribbean [@dockxDirectionalStabilizingSelection2007; @dockxNorthAmericanMonarch2004]. Despite summering and wintering in different locations, gene flow between the eastern and western monarch populations is apparently high enough to render them genetically homogeneous [@Hemstrom2022a; @tallaGenomicEvidenceGene2020;@lyonsLackGeneticDifferentiation2012].

In Southern Florida, the Caribbean, Central America, South America, and throughout most of their introduced range, however, monarchs are non-migratory [@dockxDirectionalStabilizingSelection2007;@altizerPopulationsMonarchButterflies2010;@northMonarchButterflyBiology2004]. Residency in these populations appears to be a derived trait which has arisen multiple times after the species' split from *Danaus erripus*, their closest extant relative [@Zhan2014]. The Pacific expansion of monarchs is relatively recent, with historical records indicating that they established first in Hawaii in the 1840s, likely as a result of individuals blown in during storm events that were able to survive on recently introduced milkweeds [@Zalucki2004; @Pierce2014]. Genetic evidence is consistent with an introduction to Hawaii during this time period [@Hemstrom2022a]. As described above, they then spread across the Pacific and reached Australia in approximately 1871 [@Zalucki2004], and were recorded migrating again in Australia sometime in the 1930s [see Figure 1, @James2019].

The North American migratory populations are the subject of wide-spread research on topics including but not limited to navigation and orientation [@Guerra2014; @Merlin1700; @Mouritsen10162], parasite ecology [@satterfieldMigratoryMonarchsThat2018], and life-history associated phenotypic variation [@Herman2001; @Zhan2014; @Zhan2011; @Dingle1977], and are differentiated from non-migratory individuals by a range of phenotypic variables such as reproductive diapause, wing shape, and other physiological factors [@altizerPopulationsMonarchButterflies2010; @hermanStudiesAdultReproductive1981; @browerFuelingFallMigration2006]. Note, however, that the monarch migratory syndrome is not necessarily genetically linked and can become split apart by changes in selection pressures [@Tenger-Trolander2019]. The migratory populations outside of North America, however, are not yet heavily studied, and the mechanisms enabling migration in them are not yet well understood.

@@

We assessed reproductive diapause by determining the degree of oocyte development in each of our adult female butterflies. We observed that individuals fell into two clearly defined bins: `r sum(old.data$Phenotype == "C", na.rm = TRUE)` out of `r nrow(old.data)` had fully developed, chorionated eggs (visible ridges along the exterior of the egg), while `r sum(old.data$Phenotype != "C", na.rm = TRUE)` females did not. Of these latter individuals, a few (`r sum(old.data$Phenotype == "I", na.rm = TRUE) + sum(old.data$Phenotype == "N", na.rm = TRUE)`) had no yolk in their eggs whatsoever, but most had some degree of yolking. Since vitellogenesis generally only begins to a substantial degree after eclosion in monarchs [@Pan1976], we called any individuals with more than a small amount of yolking in their eggs as reproductively mature (N = `r sum(sample.meta$likely_pheno[sample.meta$year == 2018] == "M", na.rm = TRUE)` and `r sum(sample.meta$likely_pheno[sample.meta$year == 2018] == "R", na.rm = TRUE)` for reproductively immature and mature individuals, respectively, images of egg development in each individual are available from the authors upon request). Wings and bodies from all individuals were preserved in dry coin envelops or 95% ethanol, respectively. We note that while we refer to the postponement of reproductive development under other suitable conditions as diapause, other authors refer to this process in monarchs as oligopause [@jamesOvarianDormancyDanaus1982] or quiescence, since diapause implies an extended refractory period.

@@

In order to determine if diapause is part of a generalized migratory syndrome in Australia, we also briefly re-analyzed the connection between reproductive status and wing morphology in our 2016 samples [@Freedman2018]. In order to determine if wing shape or size was correlated with reproductive status, we fit a pair of linear mixed effect models with either with wing shape or size as response variables, the number of yolked oocytes as a fixed effect, and maternal family as a random effect using the R package nlme [@Pinheiro2007].

@@

Genetic diversity levels were extremely similar between our two years ($\pi = 5.21 \times 10^{-8}$ and $5.23 \times 10^{-8}$ and $H_{o} = 5.28 \times 10^{-8}$ and $5.29 \times 10^{-8}$ for 2016 and 2018, respectively). Each year had a slight heterozygote excess ($F_{IS}$ = `r round(diversity$weighted_mean_fis[2], 3)` and `r round(diversity$weighted_mean_fis[3],3)` for 2016 and 2018, respectively), although that excess was lower if all samples were considered jointly ($F_{IS}$ = `r round(diversity$weighted_mean_fis[1], 3)`). Pairwise $F_{ST}$ between the two years was small but significant (`r round(fst$weighted_mean_fst[1], 3)`, $p =$ `r round(fst$weighted_mean_fst_p[1], 3)`). We found no significant genetic divergence between the two sample years on either the PCA or the NGSadmix results, with clustering instead driven by family, although we did find that three of our 2018 maternal families were substantially different from all other samples on our first two PC axes (Figure 2).

@@

Diapause status was strongly biased by maternal family (Fisher's Exact Test, $p < 0.001$, see Figure 3a). We identified one small region of chromosome 11 (between positions ~5,589,000 to ~5,620,000) containing many loci associated with diapause status well beyond the background level of association, including three SNPs from two different RAD-tags that had a $p$-value  $\le 5 \times 10^{-5}$ (Figure 3b). This region overlaps the gene DPOGS208560 (LOC116768596), an E3 ubiquitin ligase, and was still strongly associated with diapause if only the 2018 butterflies alone were examined. We also identified one locus with a $p$-value  $\le 5 \times 10^{-5}$ on chromosome 16, but there were no other significantly associated loci nearby.

@@

Note that it is unlikely that DPOGS208560 alone controls diapause induction under declining day-length in Australian monarchs. While our study features many *individuals*, it features only 28 maternal families and thus far fewer independant samples. Our power is therefore fairly limited, and it is entirely likely that we failed to detect many causal loci for diapause onset. This is consistent with the fact that our top associated SNP that overlaps DPOGS208560 was not perfectly associated with diapause, although that may also be due to imperfect linkage with the actual causal variants. Additional studies with larger sample sizes are therefore still needed in order to better understand the mechanisms underlying diapause induction in Australian monarch butterflies.

@@

While monarchs may be able to quickly recover from the loss of migratory behavior should conditions improve, it is unlikely that this is the case for most other migratory species of conservation concern, particularly for vertebrates. For example, if we assume the latter hypothesis (migratory variance in Australian monarchs is the result of a new mutation), it is clear that monarchs have had ample evolutionary opportunity to re-acquire migratory mutations: monarchs have a generation time of approximately seven generations per year if continuously breeding, which means that monarchs in Australia went through roughly ~ `r 7*50` generations in the 50 or so years since they were first reported on the continent, and although they likely experienced a strong bottleneck initially, they were probably at a relatively large effective size for most of that time. Since the rate at which new mutations appear in a population is proportional to both effective populations size and generation time, monarchs have had a large opportunity space for the generation of new migratory alleles.

Alternatively, it is safe to assume that if migratory variance in Australian monarchs was maintained from North America, it is clear that this would only have been possible because selection was not actively acting against that variance. This is actually quite likely: monarchs in the Pacific are not exposed to substantial annual changes in day length or average temperature, and thus do not receive their main migratory cues. Environmentally triggered migratory variation, therefore, is probably selectively nearly neutral [although more fixed migratory variance, such as wing morphology, was likely selected against, driving the observed, repeated decreases in wing size and length observed in newly non-migratory monarch populations, see @Freedman2020], and is therefore more likely to be maintained in non-migratory populations. In contrast, a species with slower generation times and smaller effective population sizes would be unlikely re-evolve lost phenotypic variance on anything approaching a similar time frame, nor would they as easily hold on to phenotypic variance in the long run under neutral conditions. Nonetheless, our findings are still optimistic for the future of migration in monarch butterflies or wherever selection on migratory alleles is merely relaxed and not becoming increasingly negative.
